Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-07-03 paint-contour-tree.tex
small changes

There are numerous algorithms for computing contour trees.
An $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, was presented by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.27\linewidth]{figs/spikey1}%
    \hspace{.35in}
    \includegraphics[width=.27\linewidth]{figs/spikey2}%
    \hspace{.4in}
    \includegraphics[width=.22\linewidth]{figs/spikeyTree}
    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
    \label{fig:relative}
\end{figure}

@@

\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are \emph{equivalent} 
if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

@@

\begin{definition} 
\label{def:int-criticalJoin}
The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

Abusing notation, for now we will just call this the join tree. (In Section~\ref{sec:extreme} we carefully handle the distinction 
between the join tree and the critical join tree.)
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. They simply form a long path. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/tent2}%
    \hspace{.9in}
    \includegraphics[width=.18\linewidth]{figs/tent1}
    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
    Note we are assuming raining was done in reverse sorted order of maxima.  
    On right, flipping the direction of rain spilling.}
    \label{fig:order}
\end{figure}

\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minimum.
Analogously, we can define maximum dominant simplicial complexes,
and both are collectively called extremum dominant.

@@

The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
of comparable vs incomparable vertices.
We arrive at the primary challenge: how to sort only those critical points that are comparable, without
constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.

@@

%\FullVer{
\section{Divide and conquer through contour surgery} \label{sec:surgery}

@@

\begin{remark}
 Note that for the terrains described in this section, the number of critical points is within a constant factor of 
 the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
\end{remark}

>>>>

There are numerous algorithms for computing contour trees.
An $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, was presented by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and its contour tree $\cC(\MM)$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.26\linewidth]{figs/spikey1}%
    \hspace{.35in}
    \includegraphics[width=.26\linewidth]{figs/spikey2}%
    \hspace{.4in}
    \includegraphics[width=.215\linewidth]{figs/spikeyTree}
    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
    \label{fig:relative}
\end{figure}

@@

\begin{definition} \label{def:equiv} Two regular contours $\psi$ and $\psi'$ are \emph{equivalent} 
if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

@@

\begin{definition} 
\label{def:int-criticalJoin}
The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points.\footnote{%
\FullVer{Later, in Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
\SubmitVer{In the full version we define the critical join tree on a certain subset of $\cV$.}
} The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

Abusing notation, for now we will just call this the join tree. (In \SubmitVer{the full version}\FullVer{Section~\ref{sec:extreme}} we carefully handle the distinction 
between the join tree and the critical join tree.)
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. They simply form a long path. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.172\linewidth]{figs/tent2}%
    \hspace{1.2in}
    \includegraphics[width=.172\linewidth]{figs/tent1}
    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
    Note we are assuming raining was done in reverse sorted order of maxima.  
    On right, flipping the direction of rain spilling.}
    \label{fig:order}
\end{figure}

\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minimum.
Analogously, we can define maximum dominant complexes,
and both are collectively called extremum dominant.

@@

The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
of comparable vs incomparable vertices.
We arrive at the main challenge: how to sort only the comparable critical points, without
constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.

@@

\FullVer{
\section{Divide and conquer through contour surgery} \label{sec:surgery}

@@

\begin{remark}
 Note that for the terrains described in this section, the number of critical points is within a constant factor of 
 the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
\end{remark}
}
